---
title: "For the article"
author: "Kine Kragl Engseth"
output: 
  html_document:
    toc: true # setting a table of content     
    toc_float: true # making the table of content floating     
    theme: flatly # setting theme     
    code_folding: show # hiding code with codefolding
editor_options: 
  chunk_output_type: console
---



# Data Wrangling

## Load the data

```{r global chunk options, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE)
```

Loading the libraries

```{r load libraries}

library(tidyverse)
library(phyloseq)
library(qiime2R) # for converting .qza from QIIME2 into phyloseq objects
library(pals) # for the color palette alphabet, used in the relative abundance plots
library(vegan) # for adonis2, used to create the PERMANOVA
library(ggvenn) # used to create venn diagram
library(gplots) # for venn() used to create venn diagram
library(ranacapa) # for ggrare() used in the rarefaction curves
```




Loading phyloseq object with the output from QIIME2

```{r import with qiime2R}

# Creating variable for datapath for easier access using file.path() later
dataPath <- "./data/run1_24_saliva_nasal_samples"


# Creating a phyloseq object from the qiime2 artifacts
physeq.obj <- qza_to_phyloseq(
  features = file.path(dataPath, "3-table.qza"), 
  tree = file.path(dataPath, "6-rooted-tree.qza"), 
  taxonomy = file.path(dataPath, "4-taxonomy.qza"),
  metadata = file.path(dataPath, "16s_paper_analysis__sample-metadata.tsv"))

```


Adding a "type" column to the metadata, to distinguish between "samples" and "control"

```{r adding new column to metadata}

# Extracting metadata from the phyloseq object
raw_metadata <- sample_data(physeq.obj)

# Making sure all sample names match with the extracted metadata
if(!all(rownames(raw_metadata) == sample_names(physeq.obj))) {
  stop("Sample names in metadata do not match sample names in phyloseq object")
}

# Adding new column, will be "control" if the source column is "pos" or "neg"
raw_metadata$type <- if_else(raw_metadata$source %in% c("pos", "neg"), "control", "sample")

# Check if the sample names are still aligned after modification
if(!all(rownames(raw_metadata) == sample_names(physeq.obj))) {
  stop("Sample names mismatch after modification")
}

# Adding the altered metadata back 
sample_data(physeq.obj) <- raw_metadata
```



## Remove possible contaminations

*Remove possible contamination ASV in negative control (water sample)*


Printing dimensions before filtering the data:

```{r}
cat("Before filtering, there are", nsamples(physeq.obj), "samples and", ntaxa(physeq.obj), "ASVs.\n")
```

ASVs with non-zero counts in the negative control, are being removed from the taxonomy table

```{r}

# Subset the phyloseq object to keep only negative control samples
neg_ctrls.obj <- subset_samples(physeq.obj, source == "neg")

# Access the OTU (or ASV) table
otu_tbl_neg_ctrls <- otu_table(neg_ctrls.obj)

# Find ASVs with non-zero values in negative controls
non_zero_neg_asvs <- apply(otu_tbl_neg_ctrls, 1, function(x) any(x > 0))

# Extract the names of these ASVs
non_zero_neg_asv_names <- names(non_zero_neg_asvs[non_zero_neg_asvs])

# Full list of taxa names
allTaxa <- taxa_names(physeq.obj)

# Removing the ASVs with counts in neg control
no_neg_Taxa <- allTaxa[!(allTaxa %in% non_zero_neg_asv_names)]

# Attaching this new taxonomy to the physeq object
pruned_physeq.obj <- prune_taxa(no_neg_Taxa, physeq.obj)
```



## Remove positive and negative controls


```{r}
# Extract sample names from the filtered phyloseq object
sample_names_no_ctrl <- subset_samples(pruned_physeq.obj, type != "control") %>% 
  sample_names()

# Pruning the sample names to only keep the samples, not the controls
pruned_physeq.obj <- prune_samples(sample_names_no_ctrl, pruned_physeq.obj)

```


## Finish cleaning up non-relevant information

Selecting only the kingdom "Bacteria" in order to remove all unassigned samples, 
as well as samples that are not bacterial (such as Archaea)

```{r}
bact_physeq.obj <- subset_taxa(pruned_physeq.obj, Kingdom == "d__Bacteria")
```

Turning the `phyloseq`-object into one into a big, big dataframe, for easier plotting

```{r}
bact_melt <- psmelt(bact_physeq.obj)
```


Printing dimensions after filtering the data:

```{r}
cat("After filtering, there are", nsamples(bact_physeq.obj), "samples and", ntaxa(bact_physeq.obj), "ASVs.\n")
```


# Methods for plots

## Observed ASV boxplot

When calculating richness, it is easy to use `estimate_richness()`. For certain richness metrics, however, the rownames will be slightly changed. Creating a function to clean up the data easily, in case this happens again.


```{r}
clean_rownames <- function(df) {
    # Check if df is a data frame
    if (!is.data.frame(df)) {
        stop("Input must be a data frame")
    }

    # Remove the leading 'X' and replace '.' with '-'
    formatted_row_names <- gsub("^X", "", rownames(df))  # Remove the 'X' at the beginning
    formatted_row_names <- gsub("\\.", "-", formatted_row_names)  # Replace '.' with '-'

    # Update the row names in the data frame
    rownames(df) <- formatted_row_names
    
    # Return the modified data frame
    return(df)
}
```

Calculating richness, and cleaning up rownames

```{r}
richness.tbl <- estimate_richness(bact_physeq.obj) %>% 
  clean_rownames()
```

 
Getting the observed ASVs

```{r}
# Extracting the column for observed
obs_richness <- richness.tbl[1] %>% 
  rownames_to_column(var = "sampleID")

# Extracting metadata from the phyloseq object to add "source"
metadata_temp <- sample_data(bact_physeq.obj)
metadata_temp$sampleID <- rownames(metadata_temp)

# Adding the metadata to this table
obs_richness.tbl <- obs_richness %>% right_join(metadata_temp)

# dropping the "type" column, as it is no longer needed
obs_richness.tbl <- obs_richness.tbl[, 1:3] 
```

Plotting observed taxa in a scatter plot

```{r}
obsASV_boxplot <- ggplot(obs_richness.tbl, aes(source, Observed, fill=source)) +
  geom_boxplot() + 
  scale_fill_manual(values=c("cornflowerblue","indianred", "turquoise")) +
  stat_summary(fun=mean, geom="point", shape=23, size=4) + # adds a diamond for the mean
  geom_jitter(shape=21, width=0.2, height=0) +
  theme_bw() + 
  ggtitle("Distribution of Observed ASVs Across Sample Locations") + 
  ylab("Observed ASVs") +
  xlab("Sample location") +
  scale_x_discrete(labels=c("Nasal aspirate", "Nasal Swab", "Saliva")) + # Renaming x-axis categories
  theme(legend.position="none") + 
  theme(plot.title = element_text(hjust = 0.5))
```


## Shannon diversity boxplot



Calculating Shannon diversity 

```{r}
# Calculating shannon values for the pruned data
shannon_values <-  estimate_richness(bact_physeq.obj, measures = "Shannon") %>% 
  rownames_to_column(var = "sampleID")

shannon_values.tbl <- shannon_values %>% right_join(metadata_temp)
shannon_values.tbl <- shannon_values.tbl[, 1:3] # dropping the "type" column, as it is no longer needed
```


Plotting Shannon diversity

```{r}
shannon_values_boxplot <- ggplot(shannon_values.tbl, aes(source, Shannon, fill=source)) +
  geom_boxplot() + 
  stat_summary(fun=mean, geom="point", shape=23, size=4) + # adds a diamond for the avg mean
  scale_fill_manual(values=c("cornflowerblue","indianred", "turquoise")) + 
  geom_jitter(shape=21, width=0.2, height=0) +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(legend.position="none") + #remove the legend as it isn't needed
  scale_x_discrete(labels=c("Nasal aspirate", "Nasal Swab", "Saliva")) + # Renaming x-axis categories
  ggtitle("Shannon diversity of samples") + 
  xlab("Sample location") +
  ylab("Shannon Diversity")
```




## Relative abundance stacked barplots


### Genus


Calculating the top 20 genera, based on relative abundance of the entire dataset

```{r}
# Agglomerating the dataset on "Genus"
bact_genus <- tax_glom(bact_physeq.obj, "Genus")
bact_genus_plot <- psmelt(bact_genus)

# Converting the values to relative abundance
physeq_genus_rel_abundance <- transform_sample_counts(bact_genus, function(x) x / sum(x))

# Creating a dataframe out of the phyloseq object, in order to plot
df_genus <- psmelt(physeq_genus_rel_abundance)

# Removing everything but the abundance and Genus
df_genus <- df_genus[, c(3, 11)]

# Finding the top 20 
top_20_genus <- df_genus %>%
  group_by(Genus) %>%
  summarise(Total_Abundance = sum(Abundance)) %>% 
  arrange(desc(Total_Abundance)) %>% 
  top_n(20)

# Finding the names of the top 20
top_20_whole_genus_names <- pull(top_20_genus, "Genus")
```

Replacing every other Genus (that is not in the top 20) with "Other".

```{r}
# Adding new column, will be "control" if the source column is "pos" or "neg"
bact_genus_plot$Genus <- if_else(bact_genus_plot$Genus %in% top_20_whole_genus_names, as.character(bact_genus_plot$Genus), "Other")

# Factorizing Genus in order to decide the order in the legend (descending abundance)
bact_genus_plot <- bact_genus_plot %>%
  mutate(Genus = factor(Genus, levels = c(top_20_whole_genus_names, "Other")))
```

Plotting the relative abundance for Genus

```{r}
# Custom labelling function for facet labels
custom_labeller <- function(variable,value){
  source.labs <- c("nasal_aspirate" = "Nasal aspirate", "nasal_swab" = "Nasal swab", "saliva" = "Saliva")
  return(source.labs[value])
}


figure_top20_genus <- ggplot(bact_genus_plot, aes(x = Sample, y = Abundance, fill = Genus)) +
  geom_col(position = "fill") +
  theme_bw() +
  facet_grid(~ bact_genus_plot$source, scales = "free", space = "free",  labeller = labeller(source = custom_labeller)) +
  #scale_fill_viridis(discrete = T) +
  #scale_fill_brewer(palette = col_vector) +
  scale_fill_manual(values=unname(alphabet())) +
  guides(fill = guide_legend(keyheight = 0.2, 
                             keywidth = 0.9,
                             ncol = 1)) +
  theme(axis.text.x = element_text(angle = 90), 
        legend.position = "right",
        legend.box = "vertical", 
        legend.margin = margin(-10, 0, 0, 0),
        plot.title = element_text(hjust = 0.5),
        axis.title.x = element_text(margin = margin(t = 5, unit = "mm"))) +
  ggtitle("Comparative Overview of Top 20 Genera by Relative Abundance") + 
  ylab("Relative abundance") + 
  xlab("Sample")

figure_top20_genus
```



### Family

Calculating top 20 families:

```{r}
bact_family <- tax_glom(bact_physeq.obj, "Family")
bact_family_plot <- psmelt(bact_family)

physeq_family_rel_abundance <- transform_sample_counts(bact_family, function(x) x / sum(x))

df_family <- psmelt(physeq_family_rel_abundance)

df_family <- df_family[, c(3, 10)]

top_20_family <- df_family %>%
  drop_na() %>% 
  group_by(Family) %>%
  summarise(Total_Abundance = sum(Abundance)) %>% 
  arrange(desc(Total_Abundance)) %>% 
  top_n(20)


top_20_whole_family_names <- pull(top_20_family, "Family")
```



```{r}

# Adding new column, will be "control" if the source column is "pos" or "neg"
bact_family_plot$Family <- if_else(bact_family_plot$Family %in% top_20_whole_family_names, as.character(bact_family_plot$Family), "Other")

# Factorizing Genus in order to decide the order in the legend
bact_family_plot <- bact_family_plot %>%
  mutate(Family = factor(Family, levels = c(top_20_whole_family_names, "Other")))


# Custom labelling function for facet labels
custom_labeller <- function(variable,value){
  source.labs <- c("nasal_aspirate" = "Nasal aspirate", "nasal_swab" = "Nasal swab", "saliva" = "Saliva")
  return(source.labs[value])
}


figure_whole20_family <- ggplot(bact_family_plot, aes(x = Sample, y = Abundance, fill = Family)) +
  geom_col(position = "fill") +
  theme_bw() +
  facet_grid(~source, scales = "free", space = "free",  labeller = labeller(source = custom_labeller)) +
  #scale_fill_viridis(discrete = T) +
  #scale_fill_brewer(palette = col_vector) +
  scale_fill_manual(values=unname(alphabet())) +
  guides(fill = guide_legend(keyheight = 0.2, 
                             keywidth = 0.9,
                             ncol = 1)) +
  theme(axis.text.x = element_text(angle = 90), 
        legend.position = "right",
        legend.box = "vertical", 
        legend.margin = margin(-10, 0, 0, 0),
        plot.title = element_text(hjust = 0.5),
        axis.title.x = element_text(margin = margin(t = 5, unit = "mm"))) +
  ggtitle("Comparative Overview of Top 20 Family by Relative Abundance") + 
  ylab("Relative abundance") + 
  xlab("Sample")

figure_whole20_family
```


## Principal Coordinates Analysis

Calculating Bray-Curtis dissimilarity and making a PCoA of our raw data, using Bray-Curtis dissimilarity.

```{r}
distance.bc <- phyloseq::distance(bact_physeq.obj, method = "bray")

ord.obj <- ordinate(bact_physeq.obj, method = "PCoA", distance = distance.bc)
```

Plotting the PCoA

```{r}
PCoA_plot <- plot_ordination(bact_physeq.obj, ord.obj, color = "source") +
  theme_bw() +
  ggtitle("PCoA of Microbial Communities by Sample Site") +
  scale_color_manual(
    name = "Sample source",
    values = c("#4daf4a", "#ff7f00", "#377eb8"),
    labels = c("Nasal Aspirate", "Nasal Swab", "Saliva"))
```

## NMDS

Not sure if we are going to include it in the paper, but I am attaching my code for the NMDS that I did, for transparency.



```{r}
set.seed(3)  # set seed for random generator
nmds.obj <- ordinate(bact_physeq.obj, method = "NMDS", distance = distance.bc)
```

```{r}
nmds_plot <- plot_ordination(bact_physeq.obj, nmds.obj, color = "source",
                      title = str_c("NMDS, stress=", format(nmds.obj$stress, digits = 3)))
nmds_plot <- nmds_plot + 
  geom_point(alpha = 0.2, size = 10) +
  scale_color_manual(
    name = "Sample source",
    values = c("#3baea0", "#f67280", "#00bbf0"),
    labels = c("Nasal Aspirate", "Nasal Swab", "Saliva")) +
  ggtitle("NMDS Visualization of Microbial Sample Sites")
```




## Venn diagram


Create a list of ASVs that occur (count > 0) in each sample

```{r}
# Subset data for each site
saliva_samples <- subset_samples(bact_physeq.obj, source == "saliva")
nasal_swab_samples <- subset_samples(bact_physeq.obj, source == "nasal_swab")
nasal_aspirate_samples <- subset_samples(bact_physeq.obj, source == "nasal_aspirate")


# SALIVA

OTU_table_saliva <- otu_table(saliva_samples)
non_zero_asvs_saliva_idx <- apply(OTU_table_saliva, 1, function(x) any(x > 0))
non_zero_asv_names_saliva <- names(non_zero_asvs_saliva_idx[non_zero_asvs_saliva_idx])

# NASAL SWAB

OTU_table_nasal_swab <- otu_table(nasal_swab_samples)
non_zero_asvs_nasal_swab_idx <- apply(OTU_table_nasal_swab, 1, function(x) any(x > 0))
non_zero_asv_names_nasal_swab <- names(non_zero_asvs_nasal_swab_idx[non_zero_asvs_nasal_swab_idx])

# NASAL ASPIRATE

OTU_table_nasal_aspirate <- otu_table(nasal_aspirate_samples)
non_zero_asvs_nasal_aspirate_idx <- apply(OTU_table_nasal_aspirate, 1, function(x) any(x > 0))
non_zero_asv_names_nasal_aspirate <- names(non_zero_asvs_nasal_aspirate_idx[non_zero_asvs_nasal_aspirate_idx])

#saliva_asvs <- get_asvs(saliva_samples)
#nasal_swab_asvs <- get_asvs(nasal_swab_samples)
#nasal_aspirate_asvs <- get_asvs(nasal_aspirate_samples)

# Create a list for Venn diagram input
venn_list <- list(saliva = non_zero_asv_names_saliva, nasal_swab = non_zero_asv_names_nasal_swab, nasal_aspirate = non_zero_asv_names_nasal_aspirate)
```


This is enough to create a venn diagram. We could create a simple one like this:

```{r}
venn(venn_list)
```

But this is a bit more pretty:

```{r}
venn_diagram <- ggvenn(
  venn_list, 
  fill_color = c("#0073C2FF", "#EFC000FF", "#CD534CFF"),
  stroke_size = 0.5, set_name_size = 5, 
  show_percentage = F
  )
```


## Rarefaction curve


```{r}
rarefaction_curve <- ggrare(bact_physeq.obj, step = 1000, color = "source", plot = FALSE, se = FALSE)

rarefaction_curve <- rarefaction_curve + 
  theme_bw() +
  ggtitle("Rarefaction Curve Analysis of ASV Richness Across Samples") +
  xlab("Sequence sample size") +
  ylab("Observed ASVs") +
  scale_color_manual(
    name = "Sample source",
    values = c("#4daf4a", "#ff7f00", "#377eb8"),
    labels = c("Nasal Aspirate", "Nasal Swab", "Saliva"))
```





# Figures


## Observed ASVs - Boxplot

```{r}
obsASV_boxplot
```


## Shannon diversity - Boxplot

```{r}
shannon_values_boxplot
```



## Principal Coordinate Analysis

```{r}
PCoA_plot
```


## Relative abundance - Genus - top20

```{r}
figure_top20_genus
```



## Relative abundance - Family - top20

```{r}
figure_whole20_family
```


## Observed ASVs - Venn diagram

```{r}
venn_diagram
```


## Rarefaction curves

```{r}
rarefaction_curve
```




# Statistical Test Results

## Kruskal-Wallis and Wilcoxon rank sum test

### Observed ASV

```{r}
kruskal.test(Observed ~ source, data = obs_richness.tbl)
```

Post-hoc with Wilcoxon rank sum test to identify the pair(s) that are statistically significant.

```{r}
pairwise.wilcox.test(obs_richness.tbl$Observed, obs_richness.tbl$source,
                 p.adjust.method = "BH")
```



### Shannon values


```{r}
kruskal.test(Shannon ~ source, data = shannon_values.tbl)
```


Post-hoc with Wilcoxon rank sum test to identify the pair(s) that are statistically significant.

```{r}
pairwise.wilcox.test(shannon_values.tbl$Shannon, shannon_values.tbl$source,
                 p.adjust.method = "BH")
```



## PERMANOVA


Using the Bray-Curtis distance from before.


```{r}
adonis2(distance.bc ~ metadata_temp$source)
```




